Article Title

### 2013-12-23 plos.tex
worked on the Methods section.

The third estimator $C_{\sf sparse}$ is based on the assumption that all correlations result from direct linear associations between a sparse set of pairs of observed cells (\figref{ 02}{C}).  This assumption is enforced by setting to zero an optimal subset of the off-diagonal elements of the precision matrix.  Therefore, $C_{\sf sparse}$ is biased towards correlation structures that arise in networks of linearly interacting pairs of neurons (\figref{ 02}{C}). The estimator takes the form
\begin{equation}\label{eq:c-sparse}
C_{\sf sparse} = S^{-1},
\end{equation}
where $S$ is a sparse matrix with a large fraction of zeros off the diagonal. The estimate has one hyperparameter that determines the sparsity (fraction of off-diagonal zeros) of $S$.

The fourth estimator $C_{\sf sparse+latent}$, combines the assumptions of the previous two.  The underlying structure arises from sparse interactions between the recorded neurons and latent factors driving their activity (\figref{02}{D}). The estimator has the form \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}{eq:c-sl}
C_{\sf sparse+latent} = (S - LL^\T)^{-1}
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a $d\times p$ matrix of factor loadings. The estimator has two hyperparameters: the number of latent units $d$ and the sparsity of $S$.

\paragraph{Simulation}
To illustrate the performance of the four regularized estimators, we constructed four model populations of $p=50$~neurons each with each matching the correlation structures of a different target estimate (\figref{02}{Row 2}).  Sample correlation matrices calculated from samples of size $n=1000$ contain substantial sampling noise (\figref{02}{Row 3}) that makes the correlation structure less obvious.

@@

In this study, we adopted \emph{negative normal log-likelihood loss}\footnote{
The loss function $\loss{C,\Sigma}$ is related to the multivariate log-likelihood function $L(\Sigma|C_{\sf 0})$ as
\begin{equation*}
    L(\Sigma|C_{\sf 0}) = -\frac p 2 \ln(2\pi) -\frac p 2 \loss{\Sigma,C_{\sf 0}}
\end{equation*}
}
defined as
\begin{equation}\label{eq:loss}
    \loss{C,\Sigma} = \frac 1 p\left[\ln \det C + \Tr(C^{-1}\Sigma)\right]
\end{equation}

@@

\paragraph{Covariance estimation in neural data}
We recorded calcium activity in dense populations of neurons in the supragranular layers in primary visual cortex of anesthetized mice using fast random-access 3D scanning two-photon microscopy \cite{Reddy:2005,Katona:2012,Cotton:2013}. We presented 300 repetitions of full-field drifting gratings with two directions of motion or 150 repetitions with five directions (Fig.~\ref{fig:03}\,A) on one side of the visual field of anesthetized mice. Groups of cells loaded with a calcium-sensitive fluorescent dye were imaged and localized in 3D space (Fig.~\ref{fig:03}\,B and E) in the visual cortex on the contralateral side to the stimulus. Using acousto-optic deflectors (AOD) to steer the laser in 3D,  we recorded the somatic calcium activity from the located cells with concurrent motion detection \cite{Ecker:2013}.  This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{03}{C}). After downsampling the signal to 20 Hz, relative firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{03}{C and D}). Only cells that produced detectable calcium activity were included in the analysis (See Methods). The average stimulus response was subtracted from each trial; the remaining signals were further downsampled into 150 ms bins to compute the noise correlation matrix (\figref{ 03}{F and G}). The durations of recordings dedicated to estimating the noise correlations ranged between 15--20 minutes.  Besides the high-repetition stimulus protocol used to estimate noise correlations, an orientation tuning stimulus protocol was used to map the orientation tuning of each cell (\figref{03}{E}). Overall, 31 imaged sites from 24 animals were included in the study.

@@

In this study, we sought to uncover the optimal low-dimensional approximation of the covariance matrix for efficient estimation of the true covariance matrices in cortical microcircuits on the temporal scale of 100--200 ms corresponding to the temporal resolution of spike rates from somatic calcium signals \cite{Cotton:2013}.  We found that networks of partial correlations were more statistically efficient than low-rank approximations of correlation matrices (Fig.\;S1 E). Furthermore, estimation became still more efficient when low-rank components were allowed to be combined with sparse partial correlations. We propose that correlation structures that are supported by the data  are more likely to reflect anatomical, causal interactions. The network of sparse partial correlations suggests interactions between specific pairs whereas the low-rank represents emergent activity patterns involving collective activations of larger circuits or global fluctuations. The revealed network of partial interactions was substantially different from that revealed by thresholding correlations (\figref{05}{C, F, H, I}). In particular, it shows that many low correlations correspond to strong negative interactions (\figref{05}{F}). Intuitively, this suggests that many low interactions are as surprising or significant considering the common correlations with other neurons. The analysis of sample correlations does not ascribe significance to correlation coefficients that are surprisingly low, based on their common correlations to other cells.

@@

In our study, we focused on estimating a specific parameter of the distribution (its covariance matrix) rather than fitting a specific parametric distribution.  Under the loss function (Eq.~\ref{eq:loss}) we selected, the covariance matrix estimation methods are equivalent to fitting a multivariate Gaussian model and selecting the best structure on the basis of cross-validated log-likelihood. However, the ranking of models was preserved under other loss functions (Fig.~\ref{supp:02}), indicating that model selection results were not strongly dependent on the assumption of gaussianity. We defend the practice of evaluating and optimizing probabilistic models with respect to their parameters that are more relevant for functional connectivity rather than maximizing their cross-validated log likelihoods, which are sensitive to all parameters. For example, model A may attain higher cross-validated log likelihood than model B due to its better ability to capture the tails of the distribution while model B better estimates the correlation matrix. We maintained that model B must be more relevant for studies of functional connectivity because correlations do reflect aspects of functional connectivity.

@@

The diagonal matrix $V_{\sf 0}=C_{\sf 0}\circ I$ contains the sample variances.  The mean sample variance is computed as the normalized matrix trace of the sample covariance matrix:  $v_{\sf 0} = \frac 1 p \Tr(C_{\sf 0})$.

In estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag}), the diagonal matrix of regularized variances $V$ is estimated by  linearly shrinking of $V_{\sf 0}$ toward the mean variance with shrinkage intensity $\alpha$:
\begin{equation}
V = V_0 + \alpha \left(v_{\sf 0} I - V_{\sf 0}\right)
\end{equation}

Then the overall estimate is computed by linearly shrinking the sample covariance matrix toward $V$ with shrinkage intensity $\lambda$:
\begin{equation}
C_{\sf diag} = C_{\sf 0} + \lambda \left[V - C_{\sf 0}\right] 
\end{equation}
The {\tt corpcor} package in the programming language R also implements this estimator \cite{Schafer:2010}, although its analytical optimization of the shrinkage intensities is based on the mean square error whereas we optimized them with respect to the loss function in Eq.~\ref{eq:loss}.

Estimator $C_{\sf factor}$ (Eq.~\ref{eq:c-factor}) is the linear shrinkage the factor analysis approximation of $C_{\sf 0}$:

@@

%\section*{Figure Legends}
\newpage
\section*{Supplementary Figures}

\setcounter{figure}{0}

\begin{figure}[!ht]
    \begin{center}
        \includegraphics{./figures/Figure-Supp01.pdf}
    \end{center}
    \caption{{\bf All-to-all performance comparisons of the sample covariance matrix and the four regularized estimators.}
}
\label{supp:01}
\end{figure}

\end{document}

>>>>

The third estimator $C_{\sf sparse}$ is based on the assumption that all correlations result from direct linear associations between a sparse set of pairs of observed cells (\figref{02}{C}).  This assumption is enforced by setting to zero an optimal subset of the off-diagonal elements of the precision matrix.  Therefore, $C_{\sf sparse}$ is biased towards correlation structures that arise in networks of linearly interacting pairs of neurons (\figref{02}{C}). The estimator takes the form
\begin{equation}\label{eq:c-sparse}
C_{\sf sparse} = S^{-1},
\end{equation}
where $S$ is a sparse matrix with a large fraction of zeros off the diagonal. The estimate has one hyperparameter that determines the sparsity (fraction of off-diagonal zeros) of $S$.

The fourth estimator $C_{\sf sparse+latent}$, combines the assumptions of the previous two.  The underlying structure arises from sparse interactions between the recorded neurons and latent factors driving their activity (\figref{02}{D}). The estimator has the form \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - LL^\T)^{-1}
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a $d\times p$ matrix of factor loadings. The estimator has two hyperparameters: the number of latent units $d$ and the sparsity of $S$.

\paragraph{Simulation}
To illustrate the performance of the four regularized estimators, we constructed four model populations of $p=50$~neurons each with each matching the correlation structures of a different target estimate (\figref{02}{\,Row 2}).  Sample correlation matrices calculated from samples of size $n=1000$ contain substantial sampling noise (\figref{02}{\,Row 3}) that makes the correlation structure less obvious.

@@

In this study, we adopted \emph{negative normal log-likelihood loss} defined as
\begin{equation}\label{eq:loss}
    \loss{C,\Sigma} = \frac 1 p\left[\ln \det C + \Tr(C^{-1}\Sigma)\right]
\end{equation}
The loss function $\loss{C,\Sigma}$ is related to the multivariate log-likelihood function $L(\Sigma|C_{\sf 0})$ as
\begin{equation}
    L(\Sigma|C_{\sf 0}) = -\frac p 2 \ln(2\pi) -\frac p 2 \loss{\Sigma,C_{\sf 0}}
\end{equation}

@@

\paragraph{Covariance estimation in neural data}
We recorded calcium activity in dense populations of neurons in the supragranular layers in primary visual cortex of anesthetized mice using fast random-access 3D scanning two-photon microscopy \cite{Reddy:2005, Katona:2012, Cotton:2013}. We presented 300 repetitions of full-field drifting gratings with two directions of motion or 150 repetitions with five directions (\figref{03}{A}) on one side of the visual field of anesthetized mice. Groups of cells loaded with a calcium-sensitive fluorescent dye were imaged and localized in 3D space (\figref{03}{B and E} in the visual cortex on the contralateral side to the stimulus. Using acousto-optic deflectors (AOD) to steer the laser in 3D,  we recorded the somatic calcium activity from the located cells with concurrent motion detection \cite{Ecker:2013}.  This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{03}{C}). After downsampling the signal to 20 Hz, relative firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{03}{C and D}). Only cells that produced detectable calcium activity were included in the analysis (See Methods). The average stimulus response was subtracted from each trial; the remaining signals were further downsampled into 150 ms bins to compute the noise correlation matrix (\figref{ 03}{F and G}). The durations of recordings dedicated to estimating the noise correlations ranged between 15--20 minutes.  Besides the high-repetition stimulus protocol used to estimate noise correlations, an orientation tuning stimulus protocol was used to map the orientation tuning of each cell (\figref{03}{E}). Overall, 31 imaged sites from 24 animals were included in the study.

@@

In this study, we sought to uncover the optimal low-dimensional approximation of the covariance matrix for efficient estimation of the true covariance matrices in cortical microcircuits on the temporal scale of 100--200 ms corresponding to the temporal resolution of spike rates from somatic calcium signals \cite{Cotton:2013}.  We found that networks of partial correlations were more statistically efficient than low-rank approximations of correlation matrices (Fig.\;S\ref{supp:01}\;E). Furthermore, estimation became still more efficient when low-rank components were allowed to be combined with sparse partial correlations. We propose that correlation structures that are supported by the data  are more likely to reflect anatomical, causal interactions. The network of sparse partial correlations suggests interactions between specific pairs whereas the low-rank represents emergent activity patterns involving collective activations of larger circuits or global fluctuations. The revealed network of partial interactions was substantially different from that revealed by thresholding correlations (\figref{05}{C, F, H, I}). In particular, it shows that many low correlations correspond to strong negative interactions (\figref{05}{F}). Intuitively, this suggests that many low interactions are as surprising or significant considering the common correlations with other neurons. The analysis of sample correlations does not ascribe significance to correlation coefficients that are surprisingly low, based on their common correlations to other cells.

@@

In our study, we focused on estimating a specific parameter of the distribution (its covariance matrix) rather than fitting a specific parametric distribution.  Under the loss function (Eq.~\ref{eq:loss}) we selected, the covariance matrix estimation methods are equivalent to fitting a multivariate Gaussian model and selecting the best structure on the basis of cross-validated log-likelihood. However, the ranking of models was preserved under other loss functions (Fig.~S\ref{fig:supp02}), indicating that model selection results were not strongly dependent on the assumption of gaussianity. We defend the practice of evaluating and optimizing probabilistic models with respect to their parameters that are more relevant for functional connectivity rather than maximizing their cross-validated log likelihoods, which are sensitive to all parameters. For example, model A may attain higher cross-validated log likelihood than model B due to its better ability to capture the tails of the distribution while model B better estimates the correlation matrix. We maintained that model B must be more relevant for studies of functional connectivity because correlations do reflect aspects of functional connectivity.

@@

The diagonal matrix $D_{\sf 0}=C_{\sf 0}\circ I$ contains the sample variances.  The mean sample variance is computed as the normalized matrix trace of the sample covariance matrix:  $v_{\sf 0} = \frac 1 p \Tr(C_{\sf 0})$.

In estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag}), the diagonal matrix of regularized variances $D$ is estimated by  linearly shrinking of $D_{\sf 0}$ toward the mean variance with shrinkage intensity $\alpha$:
\begin{equation}
D = (1-\alpha)D_0 + \alpha v_{\sf 0}I 
\end{equation} 
Then the overall estimate is computed by linearly shrinking the sample covariance matrix toward $V$ with shrinkage intensity $\lambda$ (Eq.~\ref{eq:c-diag}). The {\tt corpcor} package in the programming language R also implements this estimator \cite{Schafer:2010}, although its analytical optimization of the shrinkage intensities is based on the mean square error whereas we optimized them with respect to the loss function in Eq.~\ref{eq:loss}.

Estimator $C_{\sf factor}$ is the linear shrinkage the factor analysis approximation of $C_{\sf 0}$ (Eq.~\ref{eq:c-factor}). The target $F$ by minimizing the loss $\loss{F,C_{\sf}}$, which is equivalent to maximizing the Gaussian log-likelihood with respect to $C_{\sf 0}$. To solve this minimization problem, we used the expectation-maximization (EM) algorithm.

Estimator $C_{\sf sparse}$ (Eq.~\ref{eq:c-sparse}) finds the sparse precision matrix $S$ that minimizes the loss $\loss{S^{-1},C_{\sf 0}}$.  It is computed in two steps.  First, the zero structure of the precision matrix is obtained by minimizing the $L_1$-loss:
\begin{equation}
S = \argmin\limits_{S^\prime \succ 0} \loss{{S^\prime }^{-1},C_{\sf 0}} + \lambda \|S^\prime\|_1
\end{equation}
where $S \succ 0$ is the constraint that $S$ be a positive definite matrix and $\|S\|_1$ is the element-wise $L_1$ norm of the matrix $S$. This formulation is known as \emph{graphical lasso} \cite{Friedman:2008}. To solve this minimization problem, we modified the alternative-direction method of multipliers (ADMM) algorithm developed by \cite{Ma:2013}.  Then, the $L_1$ penalty was lifted and only the non-zero elements of $S$ where fitted to minimize the loss function $\loss{S^{-1},C_{\sf 0}}$.

Estimator $C_{\sf sparse+lowrank}$ (Eq.~\ref{eq:c-sl}) finds a precision matrix $S^*$ of dimensions $(p+d)\times(p+d)$, where $d$ is the number of latent units
\begin{equation}
S^*=
\begin{pmatrix}
S_{11} & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix}
\end{equation}
Then the covariance matrix of the observed population is 
\begin{equation}
C = \left(S_{11}-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
We require that $S_{11}$ be sparse. Rather than estimate $S_{12}$ and $S_{22}$ individually, we estimate the $p\times d$ matrix $L$ such that $S_{12}S_{22}^{-1}S_{12}=LL^\T$.

@@

%\section*{Figure Legends}
\newpage
\section*{Supplementary Figures}
\setcounter{figure}{0}
\begin{figure}[!ht]
\begin{center}
\includegraphics{./figures/Figure-Supp01.pdf}
\end{center}
\caption{{\bf All-to-all performance comparisons of the sample covariance matrix and the four regularized estimators.}
}
\label{supp:01}
\end{figure}
\end{document}
